thesis2024

### 2025-05-06 01-intro.qmd
Small improvements to the intro for cohesion with the discussion and the conclusion

# Introduction

@@

To achieve this, I will address the following research questions:

@@

The research is structured into three peer-reviewed articles (Chapters 2, 3, and 5) and one additional chapter, each addressing different aspects of chert procurement and usage. A concluding discussion integrates the findings and examines their broader implications for understanding human behaviour and adaptation over time.

-   Chapter 2, titled "Creating Frames of Reference for Chert Exploitation during the Late Pleistocene in Southwesternmost Iberia" [@belmiroCreatingFramesReference2023] offers a geological characterisation of chert sources in the Algarve region, employing macroscopic and petrographic (thin section analysis) methods. It presents a reference collection and database of local and regional chert resources, which is crucial for distinguishing between local and non-local cherts when analysing an archaeological assemblage.

-   Chapter 3, titled "Within and Beyond: Chert Procurement Patterns During the Upper Palaeolithic in Southwesternmost Iberia" [@belmiroChertProcurementPatterns2025] focuses on the analysis of chert artefacts from the UP sequence at the archaeological site of Vale Boi, using both macroscopic and petrographic (thin section analysis) methods. These artefacts are compared with geological reference collections. The chapter offers new insights into the presence of local and non-local cherts across the various UP occupations, identifies the probable sources of non-local cherts, and explores changes in the distribution of chert types over time.

-   Chapter 4, titled "Within Chert: A Multi-Technique Mineral and Geochemical Approach to the Study of Chert in Southwestern Iberia," builds on the previous chapters by applying various techniques to analyse geological and archaeological chert samples. It provides data on the mineral and chemical composition of cherts through X-Ray Diffraction (XRD), Scanning Electron Microscopy and Energy Dispersive X-Ray Spectroscopy (SEM-EDS), and portable X-Ray Fluorescence (pXRF). These methods allow for a more comprehensive characterisation of the samples and enable the use of chemical composition to refine the source attributions of archaeological cherts.

-   Chapter 5, titled "From Stone to Tool: How Raw Materials Influenced Upper Palaeolithic Technology at Vale Boi, Southwestern Iberia," combines the previously obtained data on chert types (particularly regarding the distinction between local and non-local cherts) with technological data. It explores various patterns of chert use and management.

-   Chapter 6 synthesises the results and contributions of the previous chapters and discusses the broader implications of these patterns for understanding human behaviour.

>>>>

# Introduction {#sec-chapter1}

@@

To achieve this, we address the following research questions:

@@

The research includes three main papers and a chapter, each focusing on a specific topic necessary to answer our research questions. As a result, this dissertation is structured as follows:

-   [@sec-chapter2], titled "Creating Frames of Reference for Chert Exploitation during the Late Pleistocene in Southwesternmost Iberia" [@belmiroCreatingFramesReference2023] offers a geological characterisation of chert sources in the Algarve region, employing macroscopic and petrographic (thin section analysis) methods. The paper presents a reference collection and database of local and regional chert resources, which is crucial for distinguishing between local and non-local cherts when analysing an archaeological assemblage.

-   [@sec-chapter3], titled "Within and Beyond: Chert Procurement Patterns During the Upper Palaeolithic in Southwesternmost Iberia" [@belmiroChertProcurementPatterns2025] focuses on the analysis of chert artefacts from the UP sequence at the archaeological site of Vale Boi, using both macroscopic and petrographic (thin section analysis) methods. These artefacts are compared with geological reference collections. The paper offers new insights into the presence of local and non-local cherts across the various UP occupations, identifies the probable sources of non-local cherts, and explores changes in the distribution of chert types over time.

-   [@sec-chapter4], titled "Within Chert: A Multi-Technique Mineral and Geochemical Approach to the Study of Chert in Southwestern Iberia," builds on the previous chapters by applying various techniques to analyse geological and archaeological chert samples. The chapter provides data on the mineral and chemical composition of cherts through X-Ray Diffraction (XRD), Scanning Electron Microscopy and Energy Dispersive X-Ray Spectroscopy (SEM-EDS), and portable X-Ray Fluorescence (pXRF). These methods allow for a more comprehensive characterisation of the samples and enable the use of chemical composition to refine the source attributions of archaeological cherts.

-   [@sec-chapter5], titled "From Stone to Tool: How Raw Materials Influenced Upper Palaeolithic Technology at Vale Boi, Southwestern Iberia," combines the previously obtained data on chert types (particularly regarding the distinction between local and non-local cherts) with technological data. The paper explores various patterns of chert use and management.

-   [@sec-chapter6] synthesises the results and contributions of the previous chapters and discusses the broader implications of these patterns for understanding human behaviour.

-   [@sec-chapter7] presents a small conclusion, discusses the remaining gaps and opens avenues for future research.

-   [@sec-chapter9] includes the repository links and lists of supplementary materials belonging to the articles and chapter.
